\Large Last Writer Slices and Communication Traps

### 2013-11-06 paper.tex
lots of edits.  up through the eval.  soon to be a full work-over since the "rebirth"

\ctraps is a platform for analyzing and interposing on inter-thread
communication events during a concurrent program's execution.  We build our
support for \ctraps on top of the support for last writer slicing that we
described in Section~\ref{sec:lastwriterslices}.  Using last writer slice
information, \ctraps dynamically tracks dependences representing inter-thread
communication events. \ctraps exposes communication events as {\em traps}
during an execution that {\em trap handlers} can handle to perform analysis
and interposition.  We formalize trap delivery and handling in
Section~\ref{sec:ctsoundness}.

@@

\subsubsection{\ctraps System Support}

Our base \ctraps design delivers traps on all communicating read and write
operations using compiler and runtime support.

\paragraph{Compiler Support}
Our compiler instruments all read and write operations that access potentially
escaping memory locations with a call to a {\em trap hook} function.  Trap hooks
are passed the address of the memory word involved in the access, the thread
identifier of the accessing thread, the type of operation being performed, and
the program point of the access.

\paragraph{Runtime Support}
The \ctraps runtime implements the trap hook function.  Trap hooks have two
purposes: first, to detect communication; second, to deliver traps to \ctraps
applications.  Trap hooks detect communication by looking up the \lwt entry for
the memory location being accessed.  If the thread identifier in the \lwt entry is
different from the accessing thread, then the access is a {\em communicating
memory access}.  At each communicating memory access, the runtime delivers a
trap to the accessing thread just before the memory access is allowed to
proceed.  Traps are delivered by trap hooks to {\em trap handler routines} that
implement \ctraps applications.  Trap handler routines are like signal
handlers.  They are registered with the \ctraps runtime when the program
starts.  Trap handlers may contain arbitrary application-specific code.
Figure~\ref{fig:hookapi} shows the \ctraps trap handling API.  Trap
handlers are passed the four pieces of information passed to the trap hook, as
well as the \lwt's record of the program point and thread identifier of the
last write to the involved memory location.

@@

Given a program $P'$ already instrumented with \lwt updates, the \ctraps
compiler produces $P''$. A thread $i$ in $P''$ executes a modified instruction
sequence, $P''_{i} = \{ T(p^{i}_{j}), I(p^{i}_{1}), p^{i}_{1},$ $\ldots,
T(p^{i}_{j}), I(p^{i}_{k_{i}}), p^{i}_{k_{i}} \}$, where $T(p^{i}_{j})$ is a
trap hook call.  We consider a point $Q = \{q_1, \ldots, q_{e-2}\}$ after $e-2$
instructions of a sequentially consistent execution have executed and $q_{e} =
p^{i}_{j}$ is about to execute after its trap hook $T(p^{i}_{j})$ and \lwt
update $I(p^{i}_{j})$ execute.  The trap hook $T(p^{i}_{j})$ {\em delivers} a
trap before if $LWT[m] = (t,p^{t}_{\ell w})$ and $i \ne t$.  Informally, the
trap hook delivers a trap before the instruction if the thread that last wrote
$m$ is different from the one executing the instruction.  When a trap hook,
$T(p^{t}_{j})$, delivers a trap, thread $t$ executes an arbitrary trap handler
$H = \{h^{t}_{1}, \ldots, h^{t}_{h_{n}}\}$.  Thus, after the trap, the \lwt
update (if $p^{i}_{j} = Write(m)$), and the instruction execute, the execution
reaches the new point $Q = (q_{1}, \ldots, T(q_{e}), h^{t}_{1}, \ldots,
h^{t}_{h_{n}}, I(q_{e}), q_{e})$, and continues.

We define what makes a \ctraps design sound and complete, and then discuss the
correctness of our design.  A design is {\em sound} if it only delivers a trap
when threads have communicated.  A design is {\em complete} if it always
delivers a trap when threads have communicated.  The \ctraps compiler puts a
trap hook on all read and write operations, and only on those operations.
Assuming a correct \lwt, \ctraps is correct.  The correctness of \ctraps also
relies on data-race freedom and non-overlapping memory references.  The same
arguments for \lwt correctness apply to the correctness of trap delivery and
handling.  Most importantly, data-race freedom implies that trap hooks, \lwt
updates, and accesses are all atomic.

@@

\subsection{\ctraps Applications}
\label{sec:apps}
\ctraps applications are implemented as shared object plugins ({\em i.e.}, {\tt
.so} files on Linux) that are loaded by our runtime.  They must implement the
\ctraps handler API, which requires a trap handler function and permits a
constructor, destructor,  thread constructor,  thread destructor, which allow
initialization and disposal of global and thread-local state.  A wide variety
of \ctraps applications are possible.  Several such applications were outlined
in Section~\ref{sec:background:comm}.    We implemented two applications, a
version of a debugging analysis from prior work~\cite{cci} and communication
graph collection, similar to that in DefUse~\cite{defuse}, Recon~\cite{recon},
and DMTracker~\cite{dmtracker}.

\paragraph{CCI-Prev Implementation}
CCI-Prev is a technique from CCI~\cite{cci} that computes a set of code points
that access a memory location when the previous access to the same location was
by a different thread.  We implemented a variant of CCI-Prev using \ctraps.  On
communcating read operations, the runtime updates the LWT like it normally does
on writes.  Under this policy, any operation that accesses a location that the
LWT indicates was last accessed by a different thread is recorded.  The set of
recorded accesses is the output of the analysis.  Our \ctraps implementation of
CCI-Prev took about 10 lines of code (on top of our base system).

\paragraph{Communication Graph Implementation}
Communication graphs are fundamental to several prior debugging
techniques~\cite{recon, bugaboo, defuse}.  We implement simple communication
graph collection using \ctraps.  When a trap is delivered, our implementation records
communication graph edges composed of the code point of the last writer to the
location being accessed and the code point of the trapping access.  The set of
recorded graph edges is the communication graph that we output.  Our \ctraps
implementation of communication graphs took about 50 lines of code (including
output formatting code).

\paragraph{Other Applications}
In addition to those applications, we envision that a variety of applications
will be enabled by our simple interface.  Anomaly detection and ``likely
invariant'' techniques similar to AVIO~\cite{avio}, Daikon~\cite{daikon}, and
others are a natural fit for \ctraps.  \ctraps provides the support necessary
for implementing specification checkers for specifications related to
inter-thread communication~\cite{velodrome,oshajava}.  Guided testing
tools~\cite{cuzz,chess} and bug detection tools~\cite{ctrigger} focused on
concurrency could also easily implemented as \ctraps applications.  \ctraps
also provides an interesting platform for developing new approaches to
record/replay~\cite{chimera,fdr}.  We expect these and other
applications will become more prevalent with the adoption of \ctraps'
low-overhead, deployable support for communication tracking.

Additionally, \ctraps affords the ability for {\em users} of systems to add
functionality to or extend systems.  This facet seems especially promising for
monitoring or enforcing installation-specific security or performance policies
without the need to modify applications.  While interesting, we relegate such
user-built applications to future work.

%Our redundancy analysis applies only to instrumentation on loads,
%not stores.  Eliminating instrumentation on stores may lead to stale
%information in the \lwt, which could lead to spurious traps.  Eliminating
%reads, on the other hand, can only lead to overlooked communication events,
%never false traps.

%\subsection{Design 3: \lws}
%
%Our final design, \lws, varies the compiler support used by \ctrapsfull and
%\ctrapsmm to simply eliminate instrumentation on all read operations.  With no
%instrumentation on read operations, \lws has the lowest runtime overhead, but
%the most imprecise.  However, despite not collecting information about reads,
%\lws precisely collects last writer slices: all writes are instrumented, so the
%\lwt is a sound, complete reflection of program state.

%\section{Design Justification}

%We designed \ctraps to satisfy two design goals.
%
%First, our design must be efficient enough to run in useful deployment
%environments without hardware support or sampling.  Second, our design must
%collect enough information at run time to be useful for debugging and to
%facilitate building applications.  
%
%The constraints conflict because deployment use requires minimizing
%instrumentation overheads, but collecting information requires potentially
%high-overhead instrumentation.  Our design was guided by prior work on slicing
%and concurrency debugging, and our empirical assessment of our system
%(Section~\ref{sec:eval}).  The main design challenge is deciding which memory
%dependences to track during the program's execution.  
%
%\paragraph{Tracking Most Dependences with High Overhead} 
%
%At one extreme is a design that captures all RAW, WAW, and WAR dependence
%edges.    
%%Another example is the {\em CCI-Prev} feature 
%%used by CCI~\cite{cci} tracks inter-thread RAW, WAR, WAW, and RAR dependences (and some
%%intra-thread ones) with their {\em CCI-Prev} feature. 
%Such a design collects all dependence information, which is useful for
%debugging and building analysis applications.  However, such a design falls
%short on our efficiency goal. Tracking all dependences requires work on every
%memory operation.  The additional work must update a data structure storing
%dependences and access a thread-shared structure like the \lwt that indicates
%which threads and code points are dependent.  Prior
%work~\cite{stm,coredet,raceslicing} illustrates that the overhead of this
%design point is frequently too high for use in deployment. 
%
%%Tallam, {\em et al} see overheads nearing 1000\% for server
%%programs ({\em e.g.} Apache, MySQL) with largely independent tasks.  Without
%%sampling, CCI's overheads as high as 57,000\% for some applications. In most
%%other cases, overheads are lower, but still too high for deployed use without
%%sampling.  Note that with sampling, CCI has deployable overheads, but we are
%%looking for a deployable design that does not require sampling (Constraint 1).
%
%\paragraph{Tracking Many Dependences with Moderate Overhead}
%Short of tracking all dependences, a system may make the decision to track an
%{\em ad hoc} subset of dependences,  as we discussed in
%Section~\ref{sec:background:comm}.  Such systems work well at their specific task
%({\em e.g.}, debugging a certain class of errors).  They collect enough
%dependence information to be useful for their particular application.  However,
%their usefulness is limited because tracking dependences specific to one
%application does not generalize to new applications.  
%
%One reason for tracking a subset of dependences is to reduce overheads.  Systems
%that track {\em ad hoc} dependences tend to have lower overheads than those
%that track all dependences.
%%DefUse tracks intra- and inter-thread RAR and RAW dependences; Recon tracks
%%inter-thread WAW and WAR dependences, and approximates some other dependences.
%However, these systems mostly fall short of our efficiency goal as well, often having
%overheads greater than 1000\%~\cite{defuse,recon}. Such overheads are too high for deployed
%systems.  Some systems avoid such high overheads, but require hardware
%support~\cite{avio, bugaboo, fdr} or sampling~\cite{cbi,cci} to do so, also
%failing to satisfy our design goal.
%
%\paragraph{CTraps: Tracking Select Dependences with Low Overhead}
%We make the case that our \ctraps designs satisfies our design goals.
%
%The tracking support for \ctraps is built atop last writer slices.  Last writer
%slices require tracking only inter-thread WAW dependences.  Tracking only these
%dependences does not require read operations to be instrumented.   Read
%instrumentation has high overhead because reads are more frequent than writes.
%\lws is efficient because tracking last writer slices does not need to track
%any reads.  The overhead of \ctrapsfull and \ctrapsmm is low by default because
%communication hooks (on reads) are no-ops by default.  \ctraps applications may
%perform more analysis on reads and their performance overhead scales with the
%complexity of their analysis.  We argue that such a ``pay-as-you-go'' model
%that is efficient by default is the right approach, satisfying our efficiency
%design goal.  We carefully evalute the overheads of our systems in
%Section~\ref{sec:eval}.
%
%
%Our \ctraps designs also satisfy our usefulness design goal.  \ctraps is useful for
%implementing analyses, especially those focused on concurrent program behavior.
%Inter-thread communication dependences show how values modified in one thread
%flow to operations in another thread.  This systematically selected subset of
%dependences is general enough for \ctrapsfull and \ctrapsmm to implement
%dynamic analyses related to inter-thread data flow.  Furthermore, with
%overheads low enough for deployement, \ctraps applications are useful not just
%during development, but also at runtime and in production.  \lws is also useful
%for providing important debugging information absent from current systems.  We
%describe the debugging benefits of \lws in in Section~\ref{sec:debugging}.  
%
%%\ctrapsfull and \ctrapsmm provide the ability to track reads, but they remain efficient.  The key to
%%their efficiency is the flexibility of read hooks. By default, the hooks are
%%no-ops.  CTraps {\em applications} can carefully select which dependences to
%%track, potentially increasing overhead with increases in functionality.  The
%%baseline performance of both designs is efficient, satisfying Constraint 1.
%%The flexibility provided by read and write hooks enables a variety of useful
%%applications (two of which we implement and evaluate), satisfying Constraint 2.
%

%\section{Formalism and Correctness}
%\label{sec:soundness}

%We formalize last writer slices and \ctraps.  We then discuss the soundness and
%completeness properties of our last writer slices and \ctraps designs.
%We also discuss the impact on soundness and completeness of eliminating
%redundant read instrumentation in \ctrapsmm.

%\paragraph{Preserving Write Instrumentation}
%Our system does not eliminate any write instrumentation.  Removing
%instrumentation on writes may lead to {\em false traps}.  A false trap occurs
%when a thread executes a memory access, refers to the \lwt, and the \lwt has an
%out-of-date version of the last writer information.  False traps reflect unsoundness.
%
%
%Figure~\ref{fig:unsoundness} illustrates how removing write instrumentation
%leads to unsound trap delivery.  The figure shows memory operations in two
%threads, all of which are assumed instrumented except the triangle marked
%as uninstrumented.  The actual last thread and instruction to write is shown at
%the left alongside the state of the LWT.  The figure assumes the behavior of
%\ctrapsfull, except for the single uninstrumented write operation (all of our
%designs instrument all writes).  The LWT becomes corrupted when Thread 2
%performs its uninstrumented write -- the LWT indicates the last writer was
%Thread 1, but Thread 2's uninstrumented write is the actual last writer.  When
%Thread 2 executes its next read instruction, it refers to the corrupted LWT,
%which indicates the read is communicating.  The system delivers a false trap on
%the read.  This example illustrates the risk of eliminating write
%instrumentation.  Our designs preserve instrumentation on all writes to avoid
%this problematic situation.
%
%\begin{figure}
%\centering
%\includegraphics[width=\columnwidth]{figs/unsoundness.pdf}
%\caption{\label{fig:unsoundness}Removing write instrumentation leads to unsoundness.  CTraps does not eliminate instrumentation on writes because it is unsound.}
%\end{figure}
%

%\section{Extensions to \ctraps}
%
%\subsection{Sampling}
%
%\subsection{Context-Sensitivity}

@@

We implemented last writer slicing and \ctraps.  We built our compiler support
as a plugin for GCC ({\tt gcc-4.7.0 20111112}).  We used GCC's points-to and
escape analysis support to prune instrumentation to non-escaping memory
locations.  We used GCC's dominator and post-dominator computations to
implement our redundancy analysis.  Our compiler instruments calls to {\tt
free} and {\tt delete} as writes to the pointer being deallocated.  Our
analysis handles all code compilable using this version of GCC except for a few
cases: we do not handle accesses that compile to GCC {\tt BITFIELD\_REFERENCE}
IR types because these are undocumented, we do not handle inlined assembly
instructions, and we do not handle C++ exceptions.  Note that these limitations
of our research prototype are not fundamental and the limited cases are
uncommon.

We built our runtime system from scratch in about 1000 lines of C and C++ code.
We implemented the LWT as a fixed size 2GB array of 64-bit words.  Each entry
packs a program point and a thread identifier into the 64-bit word.  By
default, program points are instruction addresses, but we also support bounded
(2-address) context-sensitivity by packing the current instruction's address
and the nearest return address into an LWT entry. When an access to a memory
location occurs, we index into the LWT with the lower bits of the location's
address.  Our prototype implementation uses a lossy resolution policy for hash
collisions.  We use this policy in our prototype because it is faster and
simpler than a chaining concurrent hash table implementation.  Collisions are
extremely rare, so this simplification is unlikely to be a problem even in a
more industrial strength implementation.

\section{Evaluation}
\label{sec:eval}
We evaluated our design along several axes.  We illustrate the debugging
benefits of last writer slices in Section~\ref{sec:eval:debugging} using case
studies conducted using real, buggy programs on real.   We evaluate our
performance overheads to demonstrate that last writer slices and \ctraps are
feasible for use in deployed systems.  We discuss the sources of \ctraps'
overhead.   We show that \ctraps is useful by implementing two existing
applications and analyzing their performance.  We characterize our
instrumentation and the precision and performance impact of our redundancy
analysis.

\subsection{Debugging with Last Writer Slices}
\label{sec:eval:debugging}
We illustrate the debugging benefits of last writer slices using three case
studies.

@@

Last writer slices provide the key piece of information to debug this failure:
the last write to the mutex before the worker tried acquiring it was when the
main thread deleted it.  A developer is likely to begin debugging by checking
the validity of the mutex.  Upon finding it is invalid ({\em i.e.}, it has
been deleted), the developer is likely to probe to determine who deleted it.
Without the information in the \lwt, answering that question is difficult
because the main thread may have executed arbitrarily far from where it
deleted the mutex.  With the \lwt, answering that question only requires the
developer to check the mutex's entry.

@@

We used last writer slices to debug an atomicity violation in MySQL-4.0.12.
Figure~\ref{fig:mysqllws} shows involved code.  Thread 1 rotates the database's
log file, marking it closed, opening a new log file, then marking it open
again.  Log rotation should be atomic, but the lack of synchronization allows
Thread 2 to run its insert during log rotation.  The insert code checks if the
log is marked closed. If it is insert is not logged, manifesting the failure.
We added an {\tt abort} statement to the non-logging case because this behavior
is a security vulnerability that the bug report describes as ``critical''.
Adding this {\tt abort} is reasonable, as the bug report describes the
condition in the {\tt else} block as being an error.

@@

We triggered the bug by running an insert query during a log rotation.  The bug
manifested, stopping execution at point {\bf C}.  As described in the bug
report, the query went unlogged and examining the core dump revealed the log
was marked closed at point {\bf B}.  However, the core dump did not reveal {\em
why} the log was marked closed there.  Looking at the last writer slice for the
{\tt log\_type} variable shows that it was written last Thread 1 at {\bf A}.
Seeing that {\tt log\_type} is accessed concurrently in inserts and
log rotations shows where synchronization needs to be added to fix the bug.

@@

Figure~\ref{fig:jdklws} shows how the bug can manifest as a failure. The string
object pointed to by {\tt sb} in Thread 1's code contains a string buffer and a
variable called {\tt count} that represents its length.  The {\tt append()}
code should atomically read {\tt sb->count} and update the string's buffer.
The failure occurs when the {\tt append()} code in Thread 1 is interleaved with
the {\tt erase()} code in Thread 2 as indicated in the figure.  The {\tt
erase()} code updates the buffer's length, making the version Thread 1 read
stale.  Thread 1 uses its stale version in the function called at point {\bf
B}, causing a crash.

@@

This situation illustrates the value of the provenenace information furnished
by last writer slices.  At the point of the failure the last writer slice for
the length variable points to a decrement of that variable in the erase
function.  The last writer slice illustrates the need for synchronization
between the append function and the erase function.  In the absence of
provenance information, the programmer would not be aware of that relationship.

@@

We conducted our evaluation on a machine running Linux 2.6.27-7, with a 2.27GHz
8-core Xeon E5520 processor with 2-way SMT ({\em i.e.}, 16 execution contexts)
and 10GB of memory.  We evaluated \ctraps using PARSEC, a set of architectural
benchmark programs~\cite{parsec} and a collection of real server and desktop
applications.

@@

The main performance result of our work is that our designs impose performance
overheads that are low enough for use in deployed systems.
Figure~\ref{fig:perfall} shows the slowdown suffered by our benchmarks due to last writer slicing
and \ctraps running with ``no-op'' trap hooks ({\em i.e.}, no
handlers), relative to the natively executing baseline.  These data reflect
our main performance findings.

>>>>

\ctraps is a framework for analyzing and interposing on inter-thread
communication events during a program's execution.  We build \ctraps on top of
last writer slicing that we described in Section~\ref{sec:lastwriterslices}.
Using last writer slice information, \ctraps dynamically tracks 
inter-thread communication events. \ctraps exposes communication
events as {\em traps} during an execution that {\em trap handlers} can handle
to perform analysis and interposition.  We formalize trap delivery and handling
in Section~\ref{sec:ctsoundness}.

@@

\subsubsection{Compiler and Runtime Support}

\ctraps delivers software traps before all read and write operations executed
by one thread that access a memory location that was last written by another
thread, {\em i.e.}, communicating memory operations.  Traps are simply function
calls made before such operations execution.  \ctraps implements traps using a
combination of compiler and runtime library support.

\paragraph{Compiler Support}
Our compiler inserts a call to the \ctraps {\em trap hook} function immediately
before all read and write operations that access potentially escaping memory
locations.  Trap hooks are passed the address of the memory location being
accessed, the thread identifier of the accessing thread, the type of operation
being performed, and the program point of the access.

\paragraph{Runtime Support}
The \ctraps runtime implements the trap hook function,  which has two purposes.
The trap hook function detects communication and delivers traps to \ctraps
applications when communication occurs.  The trap hook function detects
communication by looking up the \lwt entry for the memory location being
accessed.  If the thread identifier in the \lwt entry is different from the
accessing thread, then the trap hook function delivers a trap to the accessing
thread, just before the memory access is allowed to proceed.  To deliver a
trap, the trap hook function calls {\em trap handler routines} that are defined
in \ctraps applications.  Trap handler routines are like signal handlers.  They
are registered with the \ctraps runtime when the program starts.  Trap handlers
may contain arbitrary application-specific code.  Figure~\ref{fig:hookapi}
shows the \ctraps trap handling API.  Trap handlers are passed the four pieces
of information passed to the trap hook, as well as the \lwt's record of the
program point and thread identifier of the last write to the involved memory
location.

@@

Given a program $P'$ already instrumented with \lwt updates, the \ctraps
compiler produces $P''$. A thread $t$ in $P''$ executes a modified instruction
sequence, $P''_{t} = \{ T(p^{t}_{j}), I(p^{t}_{1}), p^{t}_{1},$ $\ldots,
T(p^{t}_{j}), I(p^{t}_{k_{t}}), p^{t}_{k_{t}} \}$, where $T(p^{t}_{j})$ is a
trap hook call.  We consider a point $Q = \{q_1, \ldots, q_{e-2}\}$ after $e-2$
instructions of an execution have executed and $q_{e} = p^{t}_{j}$ is about to
execute after its trap hook $T(p^{t}_{j})$ and \lwt update $I(p^{t}_{j})$
execute.  The trap hook $T(p^{t}_{j})$ {\em delivers} a trap before if $LWT[m]
= (t',p^{t'}_{\ell w})$ and $t \ne t'$.  Informally, the trap hook delivers a
trap before the instruction if the thread that last wrote $m$ is different from
the one executing the instruction.  When a trap hook, $T(p^{t}_{j})$, delivers
a trap, thread $t$ executes an arbitrary trap handler $H = \{h^{t}_{1}, \ldots,
h^{t}_{h_{n}}\}$.  Thus, after the trap, the \lwt update, and the instruction
execute, the execution reaches the new point $Q = (q_{1}, \ldots, T(q_{e}),
h^{t}_{1}, \ldots, h^{t}_{h_{n}}, I(q_{e}), q_{e})$, and continues.

A \ctraps implementation must be sound and complete to be correct.  A design is
{\em sound} if it only delivers a trap when threads have communicated.  A
design is {\em complete} if it always delivers a trap when threads have
communicated.  The \ctraps compiler puts a trap hook on all read and write
operations, and only on those operations.  Assuming a correct \lwt, \ctraps is
correct.  Our \ctraps implementation also does not insert any synchronization
into the program, so it comes with the same caveats related to data-race
freedom as the \lwt.  The same arguments for \lwt correctness apply to the
correctness of trap delivery and handling.  Most importantly, data-race freedom
implies that trap hooks, \lwt updates, and accesses are all atomic.

@@

\subsection{\ctraps Applications}
\label{sec:apps}
\ctraps applications are implemented as shared library plugins that are loaded
by our runtime.  They must implement the \ctraps trap handler API, which
requires a trap handler function and permits a constructor, destructor, thread
constructor, and thread destructor, which allow initialization and disposal of
global and thread-local state.  Many useful \ctraps applications are possible,
several of which were outlined in Section~\ref{sec:background:comm}.    To
demonstrate that \ctraps can implement real analyses with little work, we
implemented two applications. The first application is a version of a debugging
analysis from CCI~\cite{cci}.  The second application is a communication graph
collector, similar to DefUse~\cite{defuse}, Recon~\cite{recon}, and
DMTracker~\cite{dmtracker}.

\paragraph{CCI-Prev Implementation}
CCI-Prev is a technique from CCI~\cite{cci} that records a set of code points
that access a memory location when the previous access to the same location was
by a different thread.  We implemented a variant of CCI-Prev using \ctraps that
records such a set of code points.  In our implementation, communcating read
operations update the LWT like writes normally do.  Under this policy, any
operation that accesses a location that was last accessed by a different thread
is recorded.  The set of recorded accesses is the output of the analysis.  Our
\ctraps implementation of CCI-Prev took about 10 lines of code (on top of our
base system).

\paragraph{Communication Graph Implementation}
Communication graphs are the basis of several prior debugging
techniques~\cite{recon, bugaboo, defuse}.  We implement communication graph
collection using \ctraps.  When a trap is delivered, our implementation records
communication graph edges composed of the code point of the last writer to the
location being accessed and the code point of the trapping access.  The set of
recorded edges is the tool's output.  Our \ctraps implementation of
communication graphs took about 50 lines of code (including output formatting
code).

\paragraph{Other Applications}
In addition to those applications, we envision that a variety of applications
will be enabled by our simple interface.  Anomaly detection and ``likely
invariant'' techniques similar to AVIO~\cite{avio}, Daikon~\cite{daikon}, and
others are a natural fit for \ctraps.  \ctraps provides the support necessary
for implementing specification checkers for specifications related to
inter-thread communication~\cite{velodrome,oshajava}.  Guided testing
tools~\cite{cuzz,chess} and bug detection tools~\cite{ctrigger} focused on
concurrency could also be implemented as \ctraps applications.  \ctraps also
provides an interesting platform for developing new approaches to
record/replay~\cite{chimera,fdr}.  \ctraps is an enabling platform for
developing these or other analyses with low baseline overheads and without the
need to build an entire concurrency analysis from scratch.

@@

We implemented last writer slicing and \ctraps.  We built our compiler support
as a plugin for GCC ({\tt gcc-4.7.0 20111112}).  We used GCC's points-to and
escape analysis support to prune instrumentation to non-escaping memory
locations.  Our compiler instruments calls to {\tt free} and {\tt delete} as
writes to the pointer being deallocated.  Our analysis handles all code
compilable using this version of GCC except for a few cases: we do not handle
accesses that compile to GCC {\tt BITFIELD\_REFERENCE} IR types because these
are undocumented, we do not handle inlined assembly instructions, and we do not
handle C++ exceptions.  Note that these limitations of our research prototype
are not fundamental and the limited cases are uncommon.

We built our runtime system from scratch in about 1000 lines of C and C++ code.
We implemented the LWT as a fixed size 2GB array of 64-bit words.  Each entry
packs a program point and a thread identifier into the 64-bit word.  By
default, program points are single instruction addresses, but we also support
limited (2-address) context-sensitivity by packing the current instruction's
address and the nearest return address into a 64-bit LWT entry. When an access
to a memory location occurs, we index into the LWT with the lower bits of the
location's address.  Our prototype implementation uses a lossy resolution
policy for hash collisions.  We use this policy in our prototype because it is
faster and simpler than a chaining concurrent hash table implementation.
Collisions are extremely rare, so this simplification is unlikely to be a
problem even in a more industrial strength implementation.

We released our implementation, free and open-sourced~\cite{ctrapsonweb}, on
the web and have had contributions to our codebase from members of the GCC
development community.

\section{Evaluation}
\label{sec:eval}
We evaluated last writer slices and \ctraps along several axes.  We show that
last writer slices provide information that is useful during debugging in
Section~\ref{sec:eval:debugging} using case studies on several buggy programs.
We evaluate our performance overheads and show that last writer slices and
\ctraps are feasible for use in deployed systems for a majority of workloads we
studied.  We then characterize our instrumentation and discuss the sources of
our system's run time overhead.   We show that \ctraps is useful by
implementing two existing applications and analyzing their performance.

\subsection{Debugging with Last Writer Slices}
\label{sec:eval:debugging}
We illustrate the debugging benefits of last writer slices by using our system
to debug several real-world programs studied previously in the debugging
literature.

@@

Last writer slices provide the key piece of information to debug this failure.
When the program fails, the core dump shows that the mutex is invalid because
it has been deleted.  The mutex's last writer slice shows that the last write
to the mutex before the worker tried to acquire it was when the main thread
deleted it.  The last writer slice reveals the connection between the the two
pieces of code involved in the bug: the code that acquires the mutex and the
code that deletes the mutex.  Note that this connection is not obvious when the
program crashes because the main thread may have executed arbitrarily far from
where it deleted the mutex before the worker thread crashes.

@@

We used last writer slices to debug an atomicity violation in MySQL-4.0.12 that
is a critical security vulnerability.  Figure~\ref{fig:mysqllws} shows the
buggy code.  Thread 1 rotates the database's log file, marking it closed,
opening a new log file, then marking it open again.  Log rotation should be
atomic, but the lack of synchronization allows Thread 2 to run its insert
during log rotation.  The insert code checks if the log is marked closed. If
the log is marked closed, the insert is not logged, manifesting the failure.
We added an {\tt abort} statement to the non-logging case because this behavior
is a security vulnerability that the bug report describes as ``critical''.
Adding this {\tt abort} is reasonable, as the bug report describes the
condition in the {\tt else} block as being an error.

@@

We triggered the bug by running an insert query during a log rotation.  The bug
manifested, stopping execution at {\bf C}.  As described in the bug report, the
query went unlogged and examining the core dump revealed the log was marked
closed at point {\bf B} ({\em e.g.}, {\tt log\_type} was {\tt LOG\_CLOSED}).
However, the core dump did not reveal {\em why} the log was marked closed
there.  The last writer slice at {\bf C} for the {\tt log\_type} variable shows
that it was written last by Thread 1 at {\bf A}.  Without the last writer
slice, the relationship between the log rotation code and the insert code is
not apparent. The last writer slice reveals that the log rotation code
interacts with the insert code incorrectly via the {\tt log\_type} variable.

@@

Figure~\ref{fig:jdklws} shows how the bug can manifest as a failure. The string
object pointed to by {\tt sb} in Thread 1's code contains a string buffer and a
variable called {\tt count} that stores its length.  The {\tt append()}
code should atomically read {\tt sb->count} and update the string's buffer.
The failure occurs when the {\tt append()} code in Thread 1 is interleaved with
the {\tt erase()} code in Thread 2 as indicated in the figure.  The {\tt
erase()} code updates the buffer's length, making the version Thread 1 read
stale.  Thread 1 uses its stale version in the function called at point {\bf
B}, causing a crash.

@@

This situation illustrates the value of the provenenace information furnished
by last writer slices.  At the point of the failure the last writer slice for
the length variable points to a decrement of that variable in the erase
function.  The last writer slice illustrates the need for synchronization
between the append function and the erase function.

@@

We conducted our evaluation on a machine running Linux 2.6.27-7, with a 2.27GHz
8-core Xeon E5520 processor with 2-way SMT ({\em i.e.}, 16 execution contexts)
and 10GB of memory.  We evaluated our system using PARSEC, a set of
architectural benchmark programs~\cite{parsec} and a collection of real server
and desktop applications.

@@

The main performance result of our work is that our designs impose performance
overheads that are low enough for use in deployed systems for many of our
experimental workloads.  Figure~\ref{fig:perfall} shows the slowdown suffered
by our benchmarks due to last writer slicing and \ctraps running with ``no-op''
trap hooks ({\em i.e.}, no handlers), relative to the natively executing
baseline.  These data reflect our main performance findings.
